---
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
    css: ["style.css"]
---

class: center, middle

# Parallel programming

.course[450X]

.institution[__Stanford University__ 

Department of Political Science  
  
---  

Toby Nowacki]

---

---

# Overview

1. Measuring speed
2. Parallel programming (locally)
3. Using Farmshare: basics
4. Using Farmshare: shell scripts

---

# Measuring execution speed

* Useful for benchmarking and troubleshooting
* .fn_highlight[sys_time()] saves current computer time
* .fn_highlight[system.time({ input })] as a wrapper

```{r, cache = TRUE}
test_function <- function(x){ Sys.sleep(x)}

start <- Sys.time()
test_function(10)
end <- Sys.time()

start - end
system.time({test_function(10)})

```

---

# Parallel processing

(Pictures)

---

# Parallel processing

* In `R`, use .fn_highlight[doParallel package].
* .fn_highlight[makeCluster(n)] specifies the number of parallel processors to be set up
* .fn_highlight[registerDoParallel(cluster)] sets them up in the backend
* .fn_highlight[%dopar%] is a new operator for this kind of task

---

# A simple example

* Setting up example with one core

```{r, include = FALSE}
library(tidyverse)
```

```{r, cache = TRUE}
get_sample_dist <- function(n, x){
	sample(x, size = n, replace = TRUE) %>% mean}

rand_seq <- runif(100000, min = 1, max = 1000)
n <- 1000
rep_size <- 10:100

one_core <- system.time({map(rep_size, ~ replicate(.x, get_sample_dist(n, rand_seq)))})
```

---

# A simple example

* Setting up parallel thread example

```{r, cache = TRUE}
library(doParallel)
cl <- makeCluster(2)
registerDoParallel(cl)
two_cores <- system.time({foreach(i = rep_size, 
               .packages = "tidyverse",
           	   .combine = c) %dopar% {
	replicate(i, get_sample_dist(n, rand_seq))
	}
})
stopCluster(cl)
```

```{r include = FALSE, cache = TRUE}
cl <- makeCluster(6)
registerDoParallel(cl)
six_cores <- system.time({foreach(i = rep_size, 
               .packages = "tidyverse",
           	   .combine = c) %dopar% {
	replicate(i, get_sample_dist(n, rand_seq))
	}
})
stopCluster(cl)
```

---

# A simple example

```{r}
one_core
two_cores
six_cores
```

---

# Bootstrap example

* Bootstrapping example (taken from `doParallel` vignette)
* Parallel processing

```{r, include = FALSE}
library(doParallel)
cl <- makeCluster(6)
registerDoParallel(cl)
```

```{r, cache = TRUE}
cl <- makeCluster(6)
registerDoParallel(cl)
x <- iris[which(iris[, 5] != "setosa"), c(1,5)]
trials <- 10000
ptime <- system.time({
	r <- foreach(icount(trials), 
	             .combine=cbind) %dopar% {
		ind <- sample(100, 100, replace=TRUE)
		result1 <- glm(x[ind,2]~x[ind,1], family=binomial(logit))
		coefficients(result1)
	}
})[3]
ptime
```

---

# Bootstrap example

* Sequential processing

```{r, cachce = TRUE}
x <- iris[which(iris[, 5] != "setosa"), c(1,5)]
trials <- 10000
ptime <- system.time({
	r <- foreach(icount(trials), 
	             .combine=cbind) %do% {
		ind <- sample(100, 100, replace=TRUE)
		result1 <- glm(x[ind,2]~x[ind,1], family=binomial(logit))
		coefficients(result1)
	}
})[3]
ptime
```

---

# Farmshare: logging in

* Sherlock for high-performance computing, Farmshare for coursework
* Login occurs through `Terminal`:

```
ssh username@rice.stanford.edu
```

* Folder navigation via .fn_highlight[cd] command
* List of elements at current level via .fn_highlight[ls]

* The same commands can be used to navigate on your local machine!

---

# Farmshare: running R

* (Sherlock only:) in order to run `R` remotely on the server, need to load module:

```
ml R
```

* (Farmshare only:) no need to load module, `R` works if you call it from command line.

---

# Farmshare

* to run a whole script remotely, best to define and run a `batch` job:

```
#!/bin/sh
#SBATCH --time=20:00:00
#SBATCH --mem=20000
#SBATCH --cpus-per-task=4
#SBATCH --job-name="test_iterations"
#SBATCH --error=TestJob.%J.stderr
#SBATCH --output=TestJob.%J.out
#SBATCH --mail-user=toby.nowacki@gmail.com
#SBATCH --mail-type=ALL
#SBATCH --workdir=/home/tnowacki/strategic_voting


Rscript code/iterations_v4.R
```

---

# Farmshare

* The file needs to be submitted to the job manager, `slurm`, as follows:

```
sbatch path_to_file/file.sh
```

* Can also specify options via command line. Multiple parameters defined in helpfile:

```
sbatch -p=bigmem path_to_file/file.sh
sbatch -help
```

* Check on status of queued jobs as follows:

```
squeue -u username
```

* Cancel jobs as follows:

```
scancel jobname
```
---

