---
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
    css: ["style.css"]
---

class: center, middle

# Introduction to Tidyverse

.course[450X]

.institution[__Stanford University__ 

Department of Political Science  
  
---  

Toby Nowacki]

---

# Overview

1. Advantages of `tidyverse`
2. Introduction to the `tidyverse` universe
3. The pipe and other operators
4. `dplyr`
5. `tidyr`
6. `purrr`

---

# Why not Base R?

* Limited amount of functions, often cumbersome

```{r}
df <- data.frame(A = c(1, 2, 3), B = c("A", "B", "C"), C = c("T", "F", "F"))
df$D <- c(4, 5, 6)
df$E <- c(7, 8, 9)
df$F <- df$D - df$E
print(df)
```

* Looking for a clean, intuitive and readable way to perform multiple operations at once
* `data.table` provides a great alternative to tidyverse if you prefer more concise (and somewhat more performant) code

---

# The tidyverse Family

* .fn_highlight[dplyr]: functions for data manipulation and aggregation
	* `mutate()`, `select()`, `filter()`, `summarise()`, `arrange()`
* .fn_highlight[tidyr]: functions to bring your data into the tidy format
	* `pivot_longer()`, `pivot_wider()`, `unnest*()`, `separate()`, `unite()`
* .fn_highlight[ggplot2]: functions for data visualisation ('grammar of graphics') -- separate workshop on that!
	* `ggplot()`
* .fn_highlight[purrr]: functions for functional programming and avoiding for-loops
	* `map()`

---

# Your best friend, the pipe

* `R` comes with a bunch of standard operators: `+`, `-`, etc...
* `tidyverse` supplies the pipe: `%>%` 
* allows you to forward the output from one function straight into another.
* Works with both `tidyverse` and other functions

```{r}
out <- c("30", "40", "60")
out2 <- as.numeric(out)
mean(out2)
```

```{r message = FALSE}
library(tidyverse)
c("30", "40", "60") %>% as.numeric %>% mean
```

---

# Set intersection

* Another useful operator is `%in%`.
* Returns a logical vector with `TRUE` when element in first object is also contained in second object.

```{r}
candidates <- c("biden", "warren", "sanders", "buttigieg", "harris")
senators <- c("warren", "sanders", "romney", "gilibrand", "schumer")

candidates %in% senators
candidates[which(candidates %in% senators)]
```

---


# A classic data aggregation problem

.challenge[.challenge_title[Challenge #2.1]

Get the mean and standard deviation of flight delays by month
]


```{r}
library(nycflights13)
head(flights)
```

---

# Base `R` solution

```{r}
mean_list <- list()
sd_list <- list()
month <- unique(flights$month)
for(y in month){
	df_sub <- subset(flights, month == y)
	mean_list[y] <- mean(df_sub$dep_delay, na.rm = T)
	sd_list[y] <- sd(df_sub$dep_delay, na.rm = T)
}
unlist(mean_list)
unlist(sd_list)
```

---

# `dplyr` solution

```{r}
flights %>%
	group_by(month) %>% 
	summarise(mean = mean(dep_delay, na.rm = TRUE),
		sd = sd(dep_delay, na.rm = TRUE))

```

---

# `dplyr` functions

* .fn_highlight[select()] selects given columns from `data.frame` or `tibble`
* .fn_highlight[mutate()] adds additional columns
* .fn_highlight[filter()] selects rows that satisfy specified condition
* .fn_highlight[by_group()] carries out further commands by group
* .fn_highlight[summarise()] returns the results of aggregate functions

---

# `dplyr` example

.challenge[.challenge_title[Example #2.1]

Are longer-distance flights more delayed?
]

* Maybe yes, because delays 'rack up' and greater risk of unforeseen emergencies with 
* Maybe no, because longer flights are run by bigger airlines and are more valuable, so are therefore prioritised

---

# `dplyr` example

Let's only select the variables that we really need:

```{r}
mini_flights <- flights %>% 
	dplyr::select(month, dep_delay, distance, dest, carrier)
mini_flights %>% head
```

---

# `dplyr` example

Only keep January to March and summarise:

```{r}
agg_flights <- mini_flights %>% mutate(ln_dist = log(distance)) %>%
	filter(month %in% 1:3) %>% 
	group_by(carrier, month) %>% 
	summarise(mean_del = mean(dep_delay, na.rm = TRUE),
		mean_dist = mean(ln_dist, na.rm = TRUE))
agg_flights
```

---

# `dplyr` example

Plot the data points:
```{r Plot, fig.height = 4, fig.width = 4}
library(ggplot2)
ggplot(agg_flights, aes(x = mean_dist, y = mean_del)) +
	geom_point(aes(colour = carrier)) +
	labs(x = "log(Flight distance)", y = "Delay")
```

---

# `tidyr`

* Remember the tidy data exercise from last class?
* .fn_highlight[pivot_longer()], .fn_highlight[pivot_wider()] for recasting data frames 
* Sometimes data manipulation or visualisation is much simpler if you recast your data frame!
* Can also be combined with `dplyr` calls and pipes

---

# `tidyr` example

```{r}
data("mtcars")
mtcars %>% head
mtcars %>% 
	dplyr::select(mpg, cyl, disp) %>% 
	mutate(cartype = rownames(.)) %>%
	pivot_longer(mpg:disp) %>% head(., n = 5)
```

---

# More `tidyverse` things

* `tidyverse` functions refer to variable names directly 
    * (the dataframe is already piped in)
    * $\rightarrow$ more on quotations and quasi-quations in [Advanced R](https://adv-r.hadley.nz/)
* Refer to piped object with .fn_highlight[.]
* Refer to a range of variables (in order) with colon .fn_highlight[var1:var4]

---

# `purrr`

* writing functions avoids repetition and provides clarity, e.g.,

```{r}
ols_coefs <- function(x, y){
	solve(t(x) %*% x) %*% t(x) %*% y
}

x <- matrix(c(1, 2, 3, 5, 5, 6), nrow = 3, ncol = 2)
y <- matrix(c(8, 8.5, 0), nrow = 3, ncol = 1)
ols_coefs(x, y)
```

---

# `purrr`

* applying functions to each row or element can be done with a for-loop or apply family
    * you've probably done that already
* .fn_highlight[map()] family provides even more convenient functions

```{r}
no_list <- list(8, 16, 99)
lapply(no_list, sqrt)
```

---

# `purrr`

```{r}
map(no_list, sqrt)

```

* Doesn't seem very useful yet...
* Range of `map`-type functions in `purrr` that make life a lot easier!
* Different `map`-type functions return different object classes (e.g., vector, data.frame...)


---

# `purrr` example

[Link to GOT Example](https://jennybc.github.io/purrr-tutorial/ls01_map-name-position-shortcuts.html)

---

```{r}
library(repurrrsive)
got_chars %>% head

```

---

```{r}
library(repurrrsive)
got_chars %>% str

```

---

.challenge[.challenge_title[Challenge #2.2]

Get the names of all characters in the dataframe.
]

---

.challenge[.challenge_title[Challenge #2.2]

Get the names of all characters in the dataframe.
]

```{r}
got_chars %>% map("name")

```

---

.challenge[.challenge_title[Challenge #2.2]

Get the names of all characters in the dataframe.
]

```{r}
got_chars %>% map_chr("name")

```

---

.challenge[.challenge_title[Challenge #2.3]

Assemble a dataframe with all the key information.
]

---

.challenge[.challenge_title[Challenge #2.3]

Assemble a dataframe with all the key information.
]

```{r, message = FALSE}
library(magrittr)
got_chars %>%
	map_dfr(extract, c("name", "culture", "gender", "born", "alive"))
```


---

# Conclusion

* Base R is *sufficient*, but often not very comfortable.
* `tidyverse` provides a useful array of functions to help us out.
* The pipe, `%>%`, allows for combining multiple data wrangling operations into one workflow.
* As always, pick whatever method works best for your specific problem: `tidyverse` might not always be the best!

