---
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
    css: ["style.css"]
---

class: center, middle

# Functional programming

.course[450X]

.institution[__Stanford University__ 

Department of Political Science  
  
---  

Toby Nowacki]

---

---

# Overview

1. Why functions?
2. Common pitfalls
2. Functionals
3. Function factories
4. Recursion

---

# Why functions?

* A function is a mapping from some inputs `\(\mathbf{X}\)` to some outputs `\(\mathbf{Y}\)`.
* Whenever we carry out the same process more than once, a function is strongly recommended
* Much more convenient for both tractability and debugging
* Allow for decomposition of complex problems into smaller pieces

---

# Basic function architecture

```{r}
foo <- function(x, y){
	return(x + y)
}
```

---

# Common pitfalls (1)

Can you spot the problem?

```{r}
foo <- function(x, y){
	return(x + y)
}

item1 <- 3
item2 <- "five"
```

```
foo(item1, item2)
```
---

# Solution (1)

```{r}
foo <- function(x, y){
	stopifnot(is.numeric(x), is.numeric(y))
	return(x + y)
}
```

```
foo(item1, item2)
# Error in foo(item1, item2) : is.numeric(y) is not TRUE
# Calls: <Anonymous> ... withCallingHandlers -> withVisible -> eval -> eval -> # foo -> stopifnot
```

* An alternative is using `tryCatch()`.

---

# Common pitfall (2)

What's wrong here?

```{r}
bar <- function(x, y, z){
	out <- x + y
	return(out)
	out_two <- out + z
	return(out_two)
}

```

---

# Common pitfall (2)

What's wrong here?

```{r}
bar <- function(x, y, z){
	out <- x + y
	return(out)
	out_two <- out + z
	return(out_two)
}

bar(2, 4, 6)
```

---

# Solution (2)

```{r}
bar <- function(x, y, z){
	out <- x + y
	cat(paste0("Intermediate output: ", out))
	out_two <- out + z
	return(out_two)
}

bar(2, 4, 6)
```

---

# Functionals

* Functions can take *other functions* as arguments!
* we've seen this before in the form of `lapply` or `map`:

```{r include = FALSE, }
library(tidyverse)
```

```{r}
vec <- 2:6
map_dbl(vec, sqrt)
```

* Other functions that rely on functionals are, for example, .fn_highlight[apply], .fn_highlight[optimize], .fn_highlight[integrate]

---

# Functionals (cont'd)

* You can write your own functions with functionals:

```{r}
print_summary <- function(data, fn){
	out <- fn(data)
	return(paste0("Statistic: ", out))
}
print_summary(c(2, 4, 4), mean)
print_summary(c(2, 4, 4), max)
```

---

# Functionals (cont'd)

* But what about this?

```{r}
blob <- c(2, 4, 4, NA)
print_summary(blob, mean)
```

* Can't pass additional arguments to `mean`:
```
print_summary(blob, mean(na.rm = FALSE))
```

---

# Functionals (cont'd)

* Fortunately, there is a shortcut!
* .fn_highlight[...] lets us pass on whatever else is specified as an input argument.

```{r}
print_summary <- function(x, f, ...){
	return(f(x, ...))
}
print_summary(blob, mean, na.rm = TRUE)
```

---

# Functionals (cont'd)

* Selecting columns in dataframes is a little bit trickier.

```{r}

df <- tibble(name = c("A", "B", "C"),
			value = c(30, 16, 45))

col_summary <- function(dataframe, col_name, f, ...){
	get_col <- dataframe %>% dplyr::select(col_name)
	return(f(get_col, ...))
}
```

```
col_summary(df, value, mean)
# Error in .f(.x[[i]], ...) : object 'value' not found
```

---

# Functionals (cont'd)

* Have to rely on something called .fn_highlight[tidyeval] 
* Look up quotations and quasi-quotations!

```{r}
col_summary <- function(dataframe, col_name, f, ...){
	col_name <- enquo(col_name)
	get_col <- dataframe %>%
		summarise(out = f(!!col_name, ...))
	return(get_col)
}
col_summary(df, value, mean, na.rm = TRUE)
```

---

# Function factories

